---
title: "Project 3. Which proteins differ in the mouse model of Down syndrome "
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_section: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
library(car)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(MASS)
library(ggpubr)
library(tidyr)
library(cowplot)
library(corrplot)
library(knitr)
library(vegan)
library(factoextra)
library(plotly)
library(DESeq2)
theme_set(theme_bw())
```


In 2015, studies were conducted in a mouse model of how Down syndrome affects changes in the levels of various proteins.

We have uploaded a dataset for analysis from the [link](https://archive.ics.uci.edu/ml/datasets/Mice+Protein+Expression#)

# Dataset description 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE }
df <- read.table("Data_Cortex_Nuclear.csv", sep = ';', dec = ",", header = TRUE)
```

Analysis of the dataset showed that the experiment involved `r length(unique(gsub("_.{1,}","", df$MouseID)))` mice.

For each mouse, it was planned to carry out 15 measurements of its state, unfortunately, not all of them were performed in full. There are `r nrow(df)` measurments in dataset but only `r sum(complete.cases(df))` are full (without missing values).

We can distinguish `r length(levels(as.factor(df$class)))` different classes of mice. 

The description of these classes can be found [here](https://archive.ics.uci.edu/ml/datasets/Mice+Protein+Expression#):

Classes:

* c-CS-s: control mice, stimulated to learn, injected with saline (9 mice)
* c-CS-m: control mice, stimulated to learn, injected with memantine (10 mice)
* c-SC-s: control mice, not stimulated to learn, injected with saline (9 mice)
* c-SC-m: control mice, not stimulated to learn, injected with memantine (10 mice)

* t-CS-s: trisomy mice, stimulated to learn, injected with saline (7 mice)
* t-CS-m: trisomy mice, stimulated to learn, injected with memantine (9 mice)
* t-SC-s: trisomy mice, not stimulated to learn, injected with saline (9 mice)
* t-SC-m: trisomy mice, not stimulated to learn, injected with memantine (9 mice)

For each class in dataset there are n measurments:

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE }
by_class <- df %>% group_by(class)

kable(as.data.frame(summarise(.data=by_class, n = n())))
```

# Differences in BDNF_N product level depending on class in experiment 

We have built boxplots to visually represent differences in BDNF_N protein expression in different classes. 

```{r, echo = F, warning = F, errors = F, message=F, results="hide"}
ggplot(by_class, aes(x = class, y = BDNF_N)) + 
  geom_boxplot() + 
  ggtitle("BDNF_N level by mouse class")
```

Next, we applied one-way ANOVA to pinpoint the existence of differences.

However, first we checked the limits of its applicability. 

```{r, echo = F, warning = F, errors = F, message=F, results="hide"}
ggplot(df, aes_string(sample=df$BDNF_N)) +
  stat_qq() + stat_qq_line() +
  facet_wrap(vars(class), labeller = "label_both") + 
  ggtitle("Distribution of protein BNHF_N levels depending on the mouse class")


  
```

We used Bartlett test to check dispersion homogentity and found dispersion heteroskedasticity with p_value = `r bartlett.test(BDNF_N ~ class, df)[3]`.

However given that the number of observations in groups is close to the same, we can apply one-way ANOVA and Tukey's test in order to establish differences in groups

```{r, echo = F, warning = F, errors = F, message=F, results="hide"}
fit <- aov(BDNF_N ~ class, data=df)
summary(fit)

```

As we can see p-value is less then 0.05 so there are significant differences in mean of groups. We applied the Tukey criterion in order to find out between which groups there are differences.

```{r, echo = F, warning = F, errors = F, message=F, results="hide"}
Tukey <- TukeyHSD(fit)
plot(TukeyHSD(fit), las = 1)

```

The figure shows the differences between the group averages (Differences in mean levels of class) and their confidence intervals, calculated taking into account the control over the group error probability (95% family-wise confidence level). In 13 cases, the confidence intervals include 0, indicating no difference between the respective groups. We can conclude that there are significant differences between the other groups.

# Linear model

We built a multiple linear model, in which ERBB4_N protein level acted as a dependent variable, and all other protein levels as independent predictors 

```{r, echo = F, warning = F, errors = F, message=F}
df_protein_sub <- df[, c(2:78)]
df_protein_sub <- na.omit(df_protein_sub)
df_protein_sub_1 <- df[, c(2:78, 82)]
df_protein_sub_3 <-df[, c(1:78, 82)]
df_protein_sub_1 <- na.omit(df_protein_sub_1)
df_protein_sub_3 <- na.omit(df_protein_sub_3)
mod <- lm(ERBB4_N ~ . - pS6_N, data = df_protein_sub)
summary(mod)
```

# Model diagnostics

```{r, echo = F, warning = F, errors = F, message=F, results="hide"}
mod_diag <- fortify(mod)
gg_resid <- ggplot(data = mod_diag, aes(x = .fitted, y = .stdresid)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 2, color = "red") +
  geom_hline(yintercept = -2, color = "red") +
  labs(title = "Distribution of residues", x = "Fitted resid.", y = "Standart. resid.")

plot1 <- ggplot(mod_diag, aes(x = 1:nrow(mod_diag), y = .cooksd)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = 2, color = "red") + 
  labs(title = "Cook's distance graph", x = "Observations", y = "Cook's dist.")
# plot2 <- qqPlot(mod$residuals)
plot2 <- ggplot(mod, aes(sample=mod$residuals))+stat_qq() + labs(title = "Distribution (quantile-quantile plot)")
plot_grid(gg_resid, plot1, plot2)
```

The first graph shows that the residues are unevenly distributed. We can see dispersion heteroskedasticity.  Also here we can notice nonlinearity of relationship as not all observations are in the +/- 2 standard deviation zone. In addition, the residuals are unevenly distributed, we conclude that the variables are not independent.

Cook’s distance graph shows that there is no influential observations.

Distribution graf shows that residual’s distribution is slightly differs from normal .

To assess the presence of multicollinearity in our model, we built a correlation matrix, but visualized only significant interactions (significant level = 0.8). 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
corr_matrix <- cor(df_protein_sub)
#corrplot(corr_matrix, type="upper", title = "Correlations")
ld.vars <- attributes(alias(mod)$Complete)$dimnames[[1]] # aliased coef. Remove it
mod_1 <- lm(ERBB4_N ~  BDNF_N + pBRAF_N, data = df_protein_sub)
# summary(mod_1)
corr <- corr_matrix

corr[lower.tri(corr,diag=TRUE)] <- NA 
  #drop perfect correlations
corr[corr == 1] <- NA 
corr <- as.data.frame(as.table(corr))
  #remove the NA values from above 
  corr <- na.omit(corr) 
corr <- subset(corr, abs(Freq) > 0.77) 
  #sort by highest correlation
  corr <- corr[order(-abs(corr$Freq)),] 
mtx_corr <- reshape2::acast(corr, Var1~Var2, value.var="Freq")
  
  #plot correlations visually
  corrplot(mtx_corr, is.corr=FALSE, tl.col="black", na.label=" ")
```

As we can see, there are quite a few interactions between dependent variables.

Based on all the diagnostics of the model, we can conclude that the use of a linear model in this case is not an advantageous solution. 

We can try to optimize this model.

# Improving model quality 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE , results="hide"}
vif(mod)
which.max(vif(mod))
mod_2 <- update(mod, .~. - NR1_N)
vif(mod_2)
which.max(vif(mod_2))
mod_3 <- update(mod_2, .~. - Bcatenin_N)
vif(mod_3)
which.max(vif(mod_3))
mod_4 <- update(mod_3, .~. - TRKA_N)
which.max(vif(mod_4))
vif(mod_4)
mod_5 <- update(mod_4, .~. - ITSN1_N)
vif(mod_5)
which.max(vif(mod_5))
mod_6 <- update(mod_5, .~. - pERK_N)
vif(mod_6)
which.max(vif(mod_6))
mod_7 <- update(mod_6, .~. - pNR2B_N)
vif(mod_7)
which.max(vif(mod_7))
mod_8 <- update(mod_7, .~. - AcetylH3K9_N)
vif(mod_8)
which.max(vif(mod_8))
mod_9 <- update(mod_8, .~. - ERK_N)
vif(mod_9)
which.max(vif(mod_9))
mod_10 <- update(mod_9, .~. - PKCA_N)
vif(mod_10)
which.max(vif(mod_10))
mod_11 <- update(mod_10, .~. - ELK_N)
vif(mod_11)
which.max(vif(mod_11))

```

After removing 12 predictors from the model, with the largest VIF, multicollinearity is still observed in the model. 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
vif(mod_11)[c(1:16)]
```

Only 16 VIFs are showm.

Thus, it makes no sense to further try to optimize the model. Better to use another method.

# PCA creation

We performed principal component analisys of mice dataframe to reduce a number of dimensions.

```{r echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, results="hide"}
df_prep <- df[, c(2:78, 82)] %>% mutate(class = factor(class))

df_pca <- rda(df_protein_sub, scale = TRUE)

eigenvals(df_pca)
screeplot(df_pca, type = "lines", bstick = TRUE)

```

The graph shows the contribution of the components to the total variability. For further analysis, we will leave only the components whose contribution is greater than in Broken Stick Model.

## Ordination

After that, we built the ordination in the axes of the first two principal components 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE , results="hide"}
df_pca_base<- prcomp(df_protein_sub, scale = TRUE)
fviz_pca_ind(df_pca_base,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_protein_sub_1$class, # color by groups
             palette = "Dark2",
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Class"
             )
```

Unfortunately, grouping each class separately does not give clear visualization. Therefore, I tried to build an ordination with division into two types of mice (stimulated / non stimulated).

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE , results="hide"}
class1 <- gsub(".{2}CS.{1,}", "Non stimulated", df_protein_sub_1$class)
class1 <- gsub(".{2}SC.{1,}", "Stimulated", class1)
df_protein_sub_2 <- df
df_protein_sub_2 <- cbind(df_protein_sub_1[, -1], class1) 

fviz_pca_ind(df_pca_base,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_protein_sub_2$class1, # color by groups
             palette = "Dark2",
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Class"
             )

class1 <- gsub("t.{1,}", "Trisomy", df_protein_sub_1$class)
class1 <- gsub("c.{1,}", "Control", class1)
df_protein_sub_2 <- df
df_protein_sub_2 <- cbind(df_protein_sub_1[, -1], class1) 

fviz_pca_ind(df_pca_base,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_protein_sub_2$class1, # color by groups
             palette = "Dark2",
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Class"
             )


class1 <- gsub(".{1,}s", "Saline injected", df_protein_sub_1$class)
class1 <- gsub(".{1,}m", "Memantine injected", class1)
df_protein_sub_2 <- df
df_protein_sub_2 <- cbind(df_protein_sub_1[, -1], class1) 

fviz_pca_ind(df_pca_base,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_protein_sub_2$class1, # color by groups
             palette = "Dark2",
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Class"
             )
```

We can see that now the groups are more pronounced for stimulated and non stimulated mice. However there are almost no difference between Control and Down syndrom mice and stimulation with both saline and memantine affects them equally.

## Correlation biplot

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE , results="hide"}
biplot(df_pca, scaling = "species", display = "species")
```

We can conclude that there are a lot of proteins correlated with each other. Probably we should remove some of them from analisys.

## 3D plots for the first 3 components 

```{r,  echo=FALSE, warning=FALSE, error = FALSE}
plotly_df <- as.data.frame(df_pca$CA$u[,c(1:3)])
class <-df_protein_sub_1$class

plotly_df <- cbind(plotly_df, class)
#class(plotly_df$class)
fig <- plot_ly(plotly_df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~class, colors = c('#BF382A', '#0C4B8E'))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))

fig
```

As we see see points do not form any logical clusters.

We can try to plot graphs by grouping points in a different way.

We tried to plot graphs for trisomy / non-trisomy groups and stimulated / non-stimulated mice.

```{r,  echo=FALSE, warning=FALSE, error = FALSE}
plotly_df <- as.data.frame(df_pca$CA$u[,c(1:3)])
class1 <-df_protein_sub_1$class
class1 <- gsub("t.{1,}", "Trisomy", df_protein_sub_1$class)
class1 <- gsub("c.{1,}", "Control", class1)
plotly_df1 <- cbind(plotly_df, class1)
#class(plotly_df$class)
fig <- plot_ly(plotly_df1, x = ~PC1, y = ~PC2, z = ~PC3, color = ~class1, colors = c('#BF382A', '#0C4B8E'))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))

fig
```

```{r,  echo=FALSE, warning=FALSE, error = FALSE}
plotly_df <- as.data.frame(df_pca$CA$u[,c(1:3)])
class1 <-df_protein_sub_1$class
class1 <- gsub(".{2}CS.{1,}", "Non stimulated", df_protein_sub_1$class)
class1 <- gsub(".{2}SC.{1,}", "Stimulated", class1)
plotly_df1 <- cbind(plotly_df, class1)
#class(plotly_df$class)
fig <- plot_ly(plotly_df1, x = ~PC1, y = ~PC2, z = ~PC3, color = ~class1, colors = c('#BF382A', '#0C4B8E'))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))

fig
```

Only on the graph showing the groups of stimulated / not stimulated mice, we can see some differences in the groups. (As at 2D PCA plot)

## Percentage of variability attributable to each component 

Since the contribution of the last components is too small, only the first 15 are displayed 
```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
per_comp <- as.vector(df_pca$CA$eig)/sum(as.vector(df_pca$CA$eig))*100
per_comp <- cbind(paste0("PC", 1:76), as.data.frame(per_comp))
colnames(per_comp) <- c("PC_N", "percent")
kable(per_comp[c(1:15),])
```

# Search for differential proteins using DESeq2

We've count differences using DESeq2 for all possible conditions (together and separatly). 

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}

# Preparing data

df3 <- t(df_protein_sub)
df3<- as.data.frame(df3)
colnames(df3) <- df_protein_sub_3$MouseID

df3 <- mutate_each(df3, function(x) as.integer(x*100))

rownames(df3) <- colnames(df_protein_sub)

class1 <- gsub("t.{1,}", "Trisomy", df_protein_sub_1$class)
class1 <- gsub("c.{1,}", "Control", class1)

class2 <- gsub(".{1,}s", "Saline", df_protein_sub_1$class)
class2 <- gsub(".{1,}m", "Memantine", class2)

class3 <- gsub(".{1,}CS.{1,}", "Stimul", df_protein_sub_1$class)
class3 <- gsub(".{1,}SC.{1,}", "Non_stimul", class3)



meta <- data.frame(row.names =  df_protein_sub_3$MouseID,
                           condition = as.factor(df_protein_sub_3$class), 
                   diag = as.factor(class1),
                   stml = as.factor(class3),
                   drug = as.factor(class2))

dds_all_cond <- DESeqDataSetFromMatrix(countData = df3,
                              colData = meta,
                              design = ~ condition)

dds_diag <- DESeqDataSetFromMatrix(countData = df3,
                              colData = meta,
                              design = ~diag)

dds_drug <- DESeqDataSetFromMatrix(countData = df3,
                              colData = meta,
                              design = ~ drug)

dds_stml <- DESeqDataSetFromMatrix(countData = df3,
                              colData = meta,
                              design = ~ stml)
dds_all_cond <- DESeq(dds_all_cond)
res3 <- results(dds_all_cond)

dds_stml <- DESeq(dds_stml)
res_stml <- results(dds_stml)

dds_drug <- DESeq(dds_drug)
res_drug <- results(dds_drug)

dds_diag <- DESeq(dds_diag)
res_diag <- results(dds_diag)

p_low_res3 <- subset(res3, abs(res3$padj) < 0.05 )
p_low_res_stml <- subset(res_stml, abs(res_stml$padj) < 0.05 )
p_low_res_drug <- subset(res_drug, abs(res_drug$padj) < 0.05 )
p_low_res_diag <- subset(res_diag, abs(res_diag$padj) < 0.05 )

res <- res3[order(res3$padj),]
res_stml_ordered <- res_stml[order(res_stml$padj),]
res_drug_ordered <- res_drug[order(res_drug$padj),]
res_diag_ordered <- res_diag[order(res_diag$padj),]

head(res)

gene_list3 <- row.names(res)
gene_list_stml <- row.names(res_stml_ordered)
gene_list_drug <- row.names(res_drug_ordered)
gene_list_diag <- row.names(res_diag_ordered)
```

Here we can see DESeq2 calculated list of differrentially expressed proteins sorted by p_value (adjusted) taking into account all the classes at the same time.

Also we've prepared a gene list includes only reliably altered proteins. (it can be used in subsequent research )

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
# gene_list3
genes_cond <- data.frame(gene_list_diag[c(1:30)], gene_list_drug[c(1:30)], gene_list_stml[c(1:30)], gene_list3[c(1:30)])

# write(gene_list3, file = "gene_list3.txt")

```

In the table we can see the most changede genes per each condition. They can be used for Genquery or MSigDB.

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
colnames(genes_cond) <- c("Trisomy/Control", "Saline/Memantine", "Stimulated/Non_stimulated", "All")
kable(genes_cond)
```

We visualized the differences in expression in several proteins with the lowest p adjusted to estimate the effect of different influence at mice.

```{r,  echo=FALSE, warning=FALSE, error = FALSE, message=FALSE}
pl1 <- plotCounts(dds_all_cond, gene="SOD1_N", intgroup="condition")
pl2 <- plotCounts(dds_all_cond, gene="pERK_N", intgroup="condition")

```

